Environmental suitability throughout the late Quaternary explains population genetic diversity

### 2023-08-03 manuscript_english.tex
Updates from Overleaf

\title{Suitability prevalence area index in the late Quaternary explains genetic diversity in Tassel-eared Squirrels (\textit{Sciurus aberti})}
\author{Norma Hernandez, Angel Robles Nathan Upham}
\date{2023-06-27}

@@

Genetic variation among populations is known to exhibit biogeographic patterns in many species, but general rules of spatial genetic variation have not been established. In this paper, we establish a theoretical framework based on projecting environmental Grinellian niches back through time to relate the present geographic distribution of population genetic structure to a given species' historical evolutionary context. Thanks to the advance of next-generation sequencing technologies, as well as more accurate climate models and an enormous amount of information stored in biological collections, it is possible to implement this theoretical framework directly. We use the Tassel-earred Squirrel (\textit{Sciurus aberti}) as a model to jointly analyze spatial, environmental and genetic data and predict the historical endemic area of this species. Our results reveal that in cases of genetic isolation by
geographic distance, suitability prevalence along time corresponds to the genetic fixation
index of populations with respect to a source population. Populations closer to the historical endemic area present a higher genetic diversity and a lower \(F_{st}\) value. This empirical example fits into a theoretical framework, which allows us two advances: (i) we can now add a biogeographic explanation to the results obtained with population genetic methods; and (ii) we can generate maps of this genetic structure as predictive tools to support biodiversity conservation efforts. Overall, this work advances a perspective that integrates population genetics with historical patterns of species distribution.

@@

Current species distributions do not always correspond to their
historical distributions over evolutionarily significant periods. Often they differ substantially, such as in the case of postglacial range expansion in high latitude environments (e.g., REFS). This
is because environmental conditions are highly dynamic over time, and species
tend to track those environments, distributing where conditions are most favorable. That is, species environmental niches tend to be conserved (REF).

@@

In this work, we propose a method to find what we call the Suitability Prevalence Area
(SPA) that is the time-averaged area of highest environmental suitability for a given species, defined as locations with ≥ 90\% suitability over a historical time period. The SPA is an index with a double purpose: 1) identifying areas of greatest likelihood for the historical distribution of a target species (i.e., historical endemism); and 2) predicting the expected patterns of genetic diversity given the species' inferred history of population stability and connectivity.

@@

Subsequently, we delimited historical endemic areas to locations where
the prevalence of suitability remained at 90\% during this period.

@@

Locally, the growth rate will be related to the local population
size as long as it is within the boundary defined by \(\mathbf{N}_F\).
Therefore, the local population size as a function of the intrinsic growth rate, which depends on the change in environmental conditions over time, can be expressed as:
\(N_i(t+1) = R_0(\vec{e_i}(t) ) N_i(t)\). That is, the locally effective
population size is a function of ideal environmental conditions.

@@

The goal of this theoretical approach is not to know the shape of the
function of \(R_0\) but rather to find this local variation relating
this value of \(R_0\) and the function \(\mathbf{N}_F\) with a function
\$ S\_i(~vec\{e\}\_i)\$ of environmental suitability that can be
estimated from known methods of applied mathematics and machine learning
(for example the case of Maxent, Phillips, Anderson, and Schapire
(2006)), i.e.~:

@@

We can infer locally for each population \(N_i\) in the geography expect
to be proportional to the environmental suitability for each time
considered, For simplification in the index we select a \(N_i\)
population in the geography and infer that locally, this population is
proportional to the environmental suitability in each time considered,
i. e. \(N_{j} \propto S_{j}(\vec{e}_{j})\) (each index \(j\) is for the
generation time), as we describe above, so we can asume an harmonic mean
of suitabilities
\(\frac{1}{\tau}\sum_{j = 1}^{\tau} \frac{1}{S_j} = \frac{1}{S}\) and
substitue in the previous equation:

@@

\[
F_{i}^{st}= \frac{\beta_0}{S_i} + \beta_1
\] If we consider all sites in the available geography, we call this
harmonic mean \emph{Suitability Prevalence Area}
\(\mathbf{S} = \{S_i | S_i > k\}\) (SPA) since we expect that areas
where the \(F_{st}\) is lower are in areas where the prevalence over
time is higher, i.e.~environmental conditions remain suitable for a
significant period of time and therefore no effects of genetic drift of
populations have occurred. Conversely, it is expected that in
populations with a high fixation index value, suitability will not have
prevailed over the same time interval even though conditions currently
exist and populations are present.

@@

>>>>

\title{Environmental suitability throughout the late Quaternary explains population genetic diversity}
\author{Norma Hernandez, Angel Robles Nathan Upham}
\date{2023-06-27}

@@

Genetic variation among populations is known to exhibit biogeographic patterns in many species, but general rules of spatial genetic variation have not been established. In this paper, we establish a theoretical framework based on projecting environmental Grinellian niches back through time to relate the present geographic distribution of population genetic structure to a given species' historical evolutionary context.

Thanks to the advance of next-generation sequencing technologies, as well as more accurate climate models and an enormous amount of information stored in biological collections, it is possible to implement this theoretical framework directly. We use as a case study the Tassel-eared Squirrel (\textit{Sciurus aberti}) to jointly analyze spatial, environmental, and genetic data to predict the historical endemic area of this species. Our results reveal that in cases of genetic isolation by geographic distance, the prevalence of environmental suitability over time corresponds to the genetic fixation index of populations with respect to a source population. Populations closer to the historical endemic area present a higher genetic diversity and a lower \(F_{st}\) value.

This empirical example fits into a theoretical framework, which allows us two advances: (i) we can now add a biogeographic explanation to the results obtained with population genetic methods; and (ii) we can generate maps of this genetic structure as predictive tools to support biodiversity conservation efforts. Overall, this work advances a perspective that integrates population genetics with historical patterns of species distribution.

@@

Current species distributions do not always correspond to their
historical distributions over evolutionarily significant periods. Often they differ substantially, such as in the case of postglacial range expansion in high latitude environments (e.g., REFS). This is because environmental conditions are highly dynamic over time, and speciestend  to track those environments, distributing where conditions are most favorable. That is, species environmental niches tend to be conserved (REF).

@@

In this work, we propose a method to find an index that we call the Suitability Prevalence Area
(SPA) that is the area of highest environmental suitability for a given species over a historical time period. The SPA is an index with a double purpose: 1) identifying areas of greatest likelihood for the historical distribution of a target species (i.e., historical endemism); and 2) to explain the expected patterns of genetic diversity given the species' inferred history of population stability and connectivity.

@@

@@

Locally, the growth rate will be related to the local population
size as long as it is within the boundary defined by \(\mathbf{N}_F\).
Therefore, the local population size as a function of the intrinsic growth rate, which depends on the change in environmental conditions over time, can be expressed as:
\(N_i(t+1) = R_0(\vec{e_i}(t) ) N_i(t)\). That is, the locally population size is a function of ideal environmental conditions.

@@

The goal of this theoretical approach is not to know the shape of the
function of \(R\) but rather to find this local variation relating
this value of \(R\) and the function \(\mathbf{N}_F\) with a function
\$ S\_i(~vec\{e\}\_i)\$ of environmental suitability that can be
estimated from known methods of applied mathematics and machine learning
(for example the case of Maxent, Phillips, Anderson, and Schapire
(2006)), i.e.~:

@@

We can infer locally for each population \(N_i\) in the geography expect
to be proportional to the environmental suitability for each time
considered, For simplification in the index we select a \(N_i\)
population in the geography and infer that locally, this population is
proportional to the environmental suitability in each time considered,
i. e. \(N_{j} \propto S_{j}(\vec{e}_{j})\) (each index \(j\) is for the
generation time), as we describe above, so we can assume an harmonic mean
of suitabilities
\(\frac{1}{\tau}\sum_{j = 1}^{\tau} \frac{1}{S_j} = \frac{1}{S}\) and
substitue in the previous equation:

@@

\[
F_{i}^{st}= \frac{\beta_0}{S_i} + \beta_1
\]

If we consider all sites in the available geography, we call this
harmonic mean \emph{Suitability Prevalence Area}
\(\mathbf{S} = \{S_i | S_i > k\}\) (SPA) since we expect that areas
where the \(F_{st}\) is lower are in areas where the prevalence over
time is higher, i.e.~environmental conditions remain suitable for a
significant period of time and therefore no effects of genetic drift of
populations have occurred. Conversely, it is expected that in
populations with a high fixation index value, suitability will not have
prevailed over the same time interval even though conditions currently
exist and populations are present.

@@

Other topics to cover
- Dependency on \textit{Pinus ponderosa} -- is in doubt, P. p. arizonica in the USA, but in the Sierra Madre is a different pine species -- so there is likely more lability in the environmental niche than expected -- less of a Ponderosa specialist than assumed -- also the fungi that is associated with pines
